\vspec{} and Friends: A suite of utilities to model spectroscopic phase curves of 3D exoplanet atmospheres in the presence of stellar variability.

### 2024-07-21 src/tex/ms.tex
Revise vspec-vsm section

\section{Stellar Model}
\label{sec:star}

The \vspec{} star model is designed modularly to allow
for both simple and complex behaviors. Currently, it
is represented by a rectangular grid of points on the stellar surface,
each assigned an effective temperature. At any given time, the model computes
the surface coverage fractions of each temperature visible to the observer, accounting for the
spherical geometry, limb darkening, and any occultation by a transiting planet.

The attributes of the \texttt{Star} class describe the bulk properties of the star, including radius,
period, and the effective temperature of quiet photosphere. Herein we refer to this temperature as the
photosphere temperature to differentiate it from the temperature of spots, faculae, or other sources of variability.

\begin{figure}
    \centering
        \includegraphics[width=0.5\textwidth]{figures/surface_map_and_lc.pdf}
    \script{surface_map_and_lc.py}
    \caption{{\bf Top}: The surface map of a star with spots and faculae during a transit.
    The large circles are spots. Faculae display their 3D structure, as the hot wall is more prevalent near the limbs.
    {\bf Bottom}: An example lightcurve of a spotted star.}
    \label{fig:surface_map}
\end{figure}

\subsection{Stellar Spectra \label{subsec:spectra}}
Once the surface coverage is calculated, a composite spectrum is computed from a grid of PHOENIX stellar spectra
\citep{husser2013}.
As of \texttt{VSPEC 0.1} \TJ{fix!}, we have spectra between 2300 K and 3900 K, with steps of
100 K. Each spectrum has $\log{g} = 5$ and solar metalicity.
The raw spectra span from 0.1 to 20 \micron~ with $5 \times 10^{-6}$ \micron~ steps, however binned
versions are pre-computed for faster runtimes.

@@

On the Sun, spots have two regions: the dark, central umbra and the lighter, surrounding penumbra. A detailed review of sunspot behavior,
including sizes and lifetimes, is \citet{solanki2003}, however the specific solar values are not relevant to this description. However,
the spot area as a function of time is
\begin{equation}
    A(t) = \left\{
    \begin{array}{lr}
        A_0 e^{(t-t_0)/\tau}, & \text{if } t \leq t_0 \\
        A_0 - W(t-t_0), & \text{if } t > t_0
    \end{array}
    \right\}
\end{equation}
where $A_0$ is the maximum area reached, $t_0$ is the time of the maximum, $\tau$ is the exponential growth rate, and $W$ is the linear decay rate.

\vspec{} also takes in parameters describing the population of spots, such as the mean of the (lognormal) peak area distribution
\citep{bogdan1988}, the equilibrium surface coverage, and the distribution style.

@@

Their three-dimensional structure causes faculae to behave differently depending on their angle from disk-center. Close to the limb,
the hot wall is visible to the observer, and faculae appear as bright points. Near the center, however, the cool floor is exposed and
faculae appear dark. To consider this effect in the faculae lightcurve, we compute the fraction of the facula's normalized area -- the
area on the disk it would occupy as a flat spot -- that is occupied by each the hot wall and cool floor. This is done via numerical integral
along the radius of the spot.

@@

According to \citet{1997ApJ...484..479T}, faculae temperatures (of both the floor and wall) are dependent on the facula radius,
while depth appears to be constant. They also find that the smallest faculae have no visible floor, and that even at disk center they
appear as bright points. We parameterize the floor temperature to be
\begin{equation}
    \Delta T_{\rm eff, floor} = \left\{
    \begin{array}{lr}
    \text{Not visible}, & \text{if } r<r_{\rm min} \\
    m_{\rm floor}(r-r_{\rm min}) + \Delta T_{\rm eff, floor, 0}, & \text{if } r \ge r_{\rm min}
    \end{array}
    \right\}
\end{equation}

@@

\subsection{Flares \label{subsec:flares}}
Flares are an important source of stellar variability on short timescales. We use the {\sc xoflares}
package \citep{barclay2020} to generate realistic lightcurve shapes and assume a blackbody spectrum.
{\sc xoflares} uses a flare profile function empirically derived by \citet{davenport2016} to
describe a flare lightcurve. Temperature, $E$, FWHM, and the peak time together completely describe a flare's multiwavelength lightcurve.

Before simulating an observation, \vspec{} pre-computes a population of flares based on the frequency-energy relationship from \citet{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/granulation_lightcurve.pdf}
    \caption{
        Effect of granulation on the lightcurve of a star. This simulation assumes a mean
        10\% coverage by inter-granule region whose temperature is 300 K lower than the surrounding photosphere.
        It also assumes that the coverage value varies with a magnitude of 1\% on a timescale of 6 hours.
        }
    \script{granule_lc.py}
    \label{fig:gran_lc}
a\end{figure}

\section{Planet Model \label{sec:pl_model}}
Our model uses the Planetary Spectrum Generator\footnote{\href{https://psg.gsfc.nasa.gov/}{https://psg.gsfc.nasa.gov/}}
\citep[PSG,][]{villanueva2018} to calculate the planetary spectrum. We upload a 3D Global Circulation Model (GCM) to PSG's
Global Emission Spectra (GlobES) application to retrieve thermal and reflected spectra for various points in the planet's phase curve. \TJ{cite Vincent's paper}

\subsection{Interfacing through {\sc pypsg}}
\ref{subsec:pypsg}
\vspec{} interfaces with PSG using the {\sc pypsg} package \TJ{link}, which is essentially a Python wrapper for the PSG API system.

\TJ{Write much more}

\subsection{3D Planet models}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.

>>>>

\section{Stellar Variability Model}
\label{sec:star}

The \vspec{} stellar model is designed modularly to allow for both simple and complex behaviors. The model is available as the Python package \texttt{vspec-vsm} \TJ{(VSPEC Variable Star Model)}. It is spectral-model agnostic, meaning that it relies on another package (GridPolator, see Section \ref{sec:gridpolator}) to compute the actual variable spectra. Instead, the surface model takes a pair of latitude and longitude points as input and returns a dictionary $\{ T_{\rm eff} : f_{T_{\rm eff}} \}$ where the keys \teff~ are the effective temperatures of each surface component and the values $f_{T_{\rm eff}}$ are the fraction of the visible stellar disk with that effective temperature.

The main model class is the \texttt{Star} object, which describes bulk properties like radius, nominal \teff, and limb darkening parameters, in addition to acting as 
a container for objects representing sources of variability.

\subsection{Stellar Surface Grid}
Sources of variability that exist as major structures on the stellar surface (i.e. spots and faculae) are encoded as deviations from the nominal effective temperature on a grid of points representing the stellar surface. Each time the fractions of each surface component are calculated, the grid is initialized with constant \teff. The code then iterates through each spot and facula, replacing portions of the grid with the effective temperatures that make up each structure.

The simplest grid to build is one that is rectangular in latitude and longitude, and in which the area that each point represents decreases towards the poles. However, this can lead to unnecessary spatial oversampling at the poles, and may lead to poor resolution at the equator to keep the total number of points from being excessively high. To mitigate these problems \texttt{vspec-vsm} supports a ``grid'' generated using a Fibbonacci spiral to create a near-isotropic distribution of points. This is especially useful when simulating transits, giving optimal sampling at the stellar equator. It is also useful for resolving small structures on the surface, as the number of points needed for a given minimum resolution are greatly reduced.

\begin{figure}
    \centering
        \includegraphics[width=0.5\textwidth]{figures/surface_map_and_lc.pdf}
    \script{surface_map_and_lc.py}
    \caption{{\bf Top}: The surface map of a star with spots during a transit.
    The large elipses are spots, while the transiting planet occults a portion of the disk represented by the black circle.
    {\bf Bottom}: An example lightcurve of a spotted star.}
    \label{fig:surface_map}
\end{figure}

@@

On the Sun, spots have two regions shown if Figure \ref{fig:surface_map}: the dark, central umbra and the lighter, surrounding penumbra. A detailed review of sunspot behavior,
including sizes and lifetimes, is \citet{solanki2003}, however the specific solar values are not relevant to this description. However,
the spot area as a function of time is
\begin{equation}
    A(t) = \left\{
    \begin{array}{lr}
        A_0 e^{(t-t_0)/\tau}, & \text{if } t \leq t_0 \\
        A_0 - W(t-t_0), & \text{if } t > t_0
    \end{array}
    \right\}
\end{equation}
where $A_0$ is the maximum area reached, $t_0$ is the time of the maximum, $\tau$ is the exponential growth rate, and $W$ is the linear decay rate.

In addition to modeling the variability produced by a population of spots, \texttt{vspec-vsm} contains utilities to produce those populations given some description.
These parameters include the average spot area \citep[lognormally distributed, ][]{bogdan1988}, \teff, growth and decay rates, and the method for distributing spots on the surface. These \texttt{SpotGenerator} objects are responsible for evolving the spot population as the star ages. The number of spots created in some interval $\Delta t$ is:
\begin{equation}
    N(\Delta t) = \frac{4 \pi R_*^2 f_{\rm spot}}{A_{\rm mean}} \frac{\Delta t}{T}
\end{equation}
where $R_*$ is the stellar radius, $f_{\rm spot}$ is the fraction of the surface covered by spots in growth-decay equilibrium, $A_{\rm mean}$ is the lifetime-averaged mean spot area, and $T$ is the spot lifetime, defined as:
\begin{equation}
    T = \frac{A_0}{W} - \frac{1}{\tau} \ln{\left (\frac{A_{\rm min}}{A_0}\right )}
\end{equation}
where $A_{\rm min}$ is the area given to each spot when it is initialized.

There are also two modes to distribute spots on the star's surface. The first distributes them isotropically. The second, modeled after analysis of spots on the Sun by \citet{mandal2017}, concentrates the spot latitude around $\pm 15^{\circ}$.

@@

Their three-dimensional structure causes faculae's observational properties to change depending on their angle from disk-center. Close to the limb,
the hot wall is visible to the observer, and faculae appear as bright points. Near the center, however, the cool floor is exposed and
faculae appear dark. To consider this effect in the faculae lightcurve, we compute the fraction of the facula's normalized area -- the
area on the disk it would occupy as a flat spot -- that is occupied by each the hot wall and cool floor. This is done via numerical integral
along the radius of the spot.

@@

According to \citet{topka1997}, faculae temperatures (of both the floor and wall) are dependent on the facula radius,
while depth appears to be constant. They also find that the smallest faculae have no visible floor, and that even at disk center they
appear as bright points. We parameterize the floor temperature to be
\begin{equation}
    \Delta T_{\rm eff, floor} = \left\{
    \begin{array}{lr}
    \text{Not visible}, & \text{if } r<r_{\rm min} \\
    m_{\rm floor}(r-r_{\rm min}) + \Delta T_{\rm eff, floor, 0}, & \text{if } r \ge r_{\rm min}
    \end{array}
    \right\}
\end{equation}

@@

\subsection{Flares \label{subsec:flares}}
Flares are an important source of stellar variability on short timescales. \texttt{vspec-vsm}'s flare lightcurve model is based on the \texttt{xoflares} package \citep{barclay2020}, which itself is based on empirically derived lightcurve shapes from \citet{davenport2016}. However, \texttt{xoflares} is designed to fit lightcurves in a single spectral band, so we add additional parameters to extend it to a multiwavelength lightcurve. We completely describe a flare using its temperature, total energy, full width at half maximum (FWHM), and the time of its peak. We model a flare as a hot, optically thin region above the photosphere that produces a blackbody spectrum. In this model, the temperature is constant, and the sharp rise and fall see in the lightcurve is caused by the region's rapidly changing area. This simplification allows the total flare energy $E$ to be used as a normalization factor:
\begin{equation}
    \int_{-\infty}^{\infty}A\,dt = \frac{E}{\sigma T^4}
\end{equation}
where $A$ is the time-dependent area of the flare region, $T$ is the constant flare temperature, and $\sigma$ is the Stefan-Boltzmann constant. This allows flare lightcurves to be produced with a fixed bolometric luminosity.

Because the temperature is fixed, the relevant quantity for modeling the flux of a flare in a given integration is the integrated time-area $\int A\,dt$. These quantities are computed numerically and fed to \vspec{} along with $T$ to produce spectra with the appropriate absolute flux.

Before simulating an observation, \vspec{} asks \texttt{vspec-vsm} to pre-compute a population of flares based on the frequency-energy relationship from \citet{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f/{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/granulation_lightcurve.pdf}
    \caption{
        Effect of granulation on the lightcurve of a star. This simulation assumes a mean
        10\% coverage by inter-granule region whose temperature is 300 K lower than the surrounding photosphere.
        It also assumes that the coverage value varies with a magnitude of 1\% on a timescale of 6 hours.
        }
    \script{granule_lc.py}
    \label{fig:gran_lc}
\end{figure}

\subsection{Transits}
In the case of a transiting geometry, \texttt{vspec-vsm} is responsible for computing the properties of the portion of the stellar disk that is occulted. It does this by first projecting both the transiting planet and the visible part of the surface orthographically as they would be viewed by a distant observer. It then selects all the points that are near enough to the transit to be occulted. The code then iterates through each of those points and computes the fraction of it covered by the occulting circle using a 2D numerical integral.

\section{Stellar Spectral Grids}
\label{sec:gridpolator}

\TJ{TODO}

\section{Simple climate models}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.
